---
title: "Basis of R: Vectors"
author: "Kaung Myat Khant"
date: "`r Sys.Date()`"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r clean_env, include=FALSE}
rm(list = ls())
```


## R is a kind of object-oriented programming

- Everything is an object in R
- To create an object, we assign a value to a variable
- Use **<-** to assign the value. Shortcut is **Alt+-**
- To see the object, just mention its name

```{r assignment, include=TRUE}
x <- 15 ; x
a <- "space" ; a
```

## Vectors

- More than one values can be assigned to an object by concatenate **c()**
```{r concatenate, include=TRUE}
concat <- c(15,20,25,30) ; concat
```
- And we get a **numeric vector** of length **4**
- A vector is a fundamental data structure storing same type of elements

## Common types of vectors
- Numeric
- Logical
- Character
- Factor

## Common types of vectors

```{r vector, include=TRUE}
numeric <- c(1,2,3,4) ; numeric
logical <- c(TRUE, FALSE, FALSE, TRUE); logical
character <- c("a","b","c","d"); character
```

## Common types of vector  
We usually prefer factors to character strings.
```{r factor, include=TRUE}
factor <- factor(rep(c(0,1),2),levels = c(0,1), labels = c("D-","D+"), ordered =T); factor
```

## We can see the type of vector by using **str()** 
```{r str, include=TRUE}
str(numeric); str(logical); str(character)
str(factor)
```

## Creating vectors  
- We can create vectors by using the following methods:  
  1. Concatenating **c()**
  2. Sequencing **":"**, **seq()**
  3. repeating **rep()**

## Creating vectors
*Concatenating* coerce the vector elements into single data type
```{r concat, include=TRUE}
concat <- c("a","b","c",1:3, TRUE) ;  str(concat)
```

## Creating vectors  
*Sequencing* can be use to create numeric vector
```{r seq, include=TRUE}
simple_seq <- 1:10; simple_seq
complex_seq_increment <- seq(from = 0.5, to = 5, by = 0.5) ; complex_seq_increment
```

## Creating vectors
*Repeating* can be use for both character and numeric data
```{r rep, include=TRUE}
rep_cha <- rep(c("a","b"), times=5) ; rep_cha
rep_num <- rep(1:2, each=5) ; rep_num
```

## Converting vector types
- We can convert vectors from one data type to another.
  1. character using **as.character()**
  2. numeric using **as.numeric()**
  3. logical using **as.logical()**
  4. factor using **as.factor()**
- We should note that **coercion** occurs when we convert from one data type to another.

## Coercion
- *logical* data can be coerced into *numeric*, *character* or *factor*
```{r logi_coerce, include=TRUE}
as.numeric(logical)
as.character(logical)
as.factor(logical)
```

## Coercion
- When *numerical* data is coerced into *logical*, 
  - **zero** is converted to **FALSE** 
  - other numbers are converted to **TRUE**
```{r num_coerce, include=TRUE}
as.logical(numeric)
as.logical(c(-1,-2, 0, 1,1.5))
```

## Coercion
- When *character* data is coerced into *logical*, 
  - **"TRUE/T"** is converted to **TRUE** regardless of the case of the letter
  - same for **"FALSE/F"**
  - other values are coerced into **NA** (missing value)
```{r cha_coerce, include=TRUE}
as.logical(c("TRUE","true","FALSE","false","t","T","f","F"))
```

## Coercion
- A *numeric* vector can be converted into *character*
```{r num_char, include=TRUE}
as.character(1:10)
```

## Coercion
- Only *character* vector with *numeric elements* can be converted into *numeric*
- Other values are coerced into *NA*
```{r char_num, include=TRUE}
as.numeric(c("0.5","1","-1.5","a","TRUE","1/2","(3-1)"))
```

## Coercion Power
- **Character** > **Numeric** > **Logical** 
- **NA** is a logical data by default
```{r coercion, include=TRUE}
c("a",1,TRUE,NA)
c(1,TRUE,NA)
c(TRUE,NA)
```

## Hierchy of R's Objects
### Vector object
  - **Atomic vectors**
    - *logical*
    - *numeric*
      - integer
      - double
    - *character*
  - **List** 
  
### NULL object

## Why coercion is important?
- Let's see the numeric atomic vectors
  - double (default)
  - integer
- Doubles are approximations
- Integers are exacts
- We can get the atomic type by using **typeof()**

## Why coercion is important?
There can be approximation errors in some operations
```{r coercion_error, include=TRUE}
sqrt(2)^2 ; typeof(sqrt(2)^2)
sqrt(2)^2 - 2 
sqrt(2)^2 == 2
```

---  
```{r coerc_err, include=TRUE}
dplyr::near(sqrt(2)^2,2)
typeof(2)
typeof(2L)
2 == 2L
```

## Why coercion is important?
- Special values
  - Integers have one special value: **NA**
  - Doubles have four special values: 
    - **NA**
    - **-Inf**
    - **NaN** 
    - **Inf**

## Special values
```{r special_value, include=TRUE}
NA+5; mean(rep(c(1,NA), each = 100),na.rm = F)
c(-1,0,1)/0
-Inf + Inf
```

## NA types
```{r na, include=TRUE}
typeof(NA)
typeof(NA_integer_)
typeof(NA_real_)
typeof(NA_character_)
```

